---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: 'Exercise 5: Applied Data Analysis with R'
author: "Ulrich Matter"
date: "06/12/2018"
output:
  html_document:
    highlight: tango
    theme: cerulean
    mathjax: "http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../../style/notes.css
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
```



## Z-Scores and the Standard Normal Distribution

```{r}
# The Z-Score formula in R
# define the parameter values
X <- 10
mu <- 12
sigma <- 2
# compute the z-score
z <- (X - mu) / sigma
z

# Plot the Standard Normal Distribution
plot(function(x) dnorm(x), -4, 4, main = "Normal density")

# Get the area under the curve (probability of observing a value of a certain size)
pnorm(-1)
pnorm(-2)
pnorm(-1) - pnorm(-2)

```


## Standard Errors

```{r}
# forumla for standard error
# define parameter values
s <- 20
n <- 100
# compute the standard error (of the mean)
se <- s / sqrt(n)
se

# write your own standard error (of the mean) function
se <- function(x) {
     s <- sd(x)
     n <- length(x)
     se <- s / sqrt(n)
     
     return(se)
}

# draw a random sample of size 100 and compute the mean and its estimated standard error
mysample <- rnorm(100)
mean(mysample)
se(mysample)

# repeat this but this time with a larger sample
mysample <- rnorm(1000)
mean(mysample)
se(mysample)


```




## Hypothesis Testing: the T-Statistic

### Reproduce the example from the presentation
```{r}
# define parameters
mu <- 39000
sample_mean <- 37000
sample_sd <- 6150
n <- 100

# calculate the  standard error of the sample mean
se <- sample_sd / sqrt(n)

# compute the t-statistic (and compare it with the critical value)
t <- (sample_mean - mu) / se

# look up the p-value 
# (the fraction of the mass under the standard normal distribution)
2*pnorm(-abs(t))

```

### Extended Example
```{r}
# I) Compute the t-value step by step with our own implementation while controlling the properties of the distribution we are drawing a sample form

# define size of sample
n <- 100
# draw the random sample from a normal distribution with mean 10 and sd 2
sample <- rnorm(n, mean = 10, sd = 2)

# compute the sample mean
sample_mean <- mean(sample)
# compute the sample sd
sample_sd <- sd(sample)
# estimated standard error of the mean
mean_se <- sample_sd/sqrt(length(sample))

# compute the t-statistic for the null hypothesis: H0: mu = 9
t <- (sample_mean - 10) / mean_se
t 

# get the p value
2*pnorm(-abs(t))


# II) Apply the R-function t.test to do the same! 
t.test(sample, mu = 10)

```



## Exercises
**1. Got to <http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html>.**
**2. Pick a data set that interests you, load it with the data() function.**
```{r}
# yes, I choose the swiss data set...
data(swiss)
# have a look at what we are dealing with here
str(swiss)
# have a look at the first few lines
head(swiss)
# get more info about the variables in that data set
?swiss
```

**3. Pose an empirical question you want to answer with that data set.**
In the case of the `swiss` data set, for example, is the average fertility in the entire population 85?

**4. Formulate a null-hypothesis and test it with the techniques learned today.**
H0: mu = 85. This requires a t-statistic of the Fertility mean. We can compute this statistic either with the formulas and R-functions discussed on Day 2. Or we directly compute it with t.test() provided by R. The following code demonstrates both approaches step by step.

### I. Own implementation (with a detailed explanation)
```{r}

# a) compute sample mean and sample standard deviation, record how many observations we are having
# in our sample, define the population mean (that you want to test for)
sample_mean <- mean(swiss$Fertility)
sample_sd <- sd(swiss$Fertility)
n <- length(swiss$Fertility) # alternatively use nrow(swiss)
mu <- 85

# b) compute the (estimate of the) sample mean standard error
se <- sample_sd / sqrt(n)

# c) compute the t-statistic
t <- (sample_mean - mu) / se
t

# d) check what p-value is associated with that t-statistic 
# i.e., check what fraction of the standard normal distribution has an at least as extreme value as
# the t value we computed.
pval <- 2*pnorm(-abs(t))
pval
```

Note how we get from the t-value to the p-value here: `2*pnorm(-abs(t))`. This short line of code involves a lot. Let's have a look at specific aspects. Recall the standard normal distribution function which is implementd in `pnorm(x)`. This function returns the fraction of the mass under the standard normal distribution curve that is smaller than or equal to `x` (i.e., the probability of observing a value that is equal to or smaller than `x`). Lets have a closer look in R:

```{r}
# the probability of observing a value at least as small as -1 in a standard normal distribution 
pnorm(-1)
# or in percent
pnorm(-1) * 100
```

Recall that we are interested in the probability of observing a value at least as *extreme* as `t`. Thus we need to consider the absolute value of `t`, `|t|`, and want to know at once what the probability of observing a value at least as small as `-t` *or* at least as big as `t` (therefore, the `abs(t)` which returns the absolute value). 

```{r}
# demonstration of concept of absolute value
abs(-1)
abs(1)

# in our example of t
abs(t)

```

Recall that `pnorm(x)` returns by default the probability of observing a value that is at least as small as `x` therefore the `-abs(t)`. Finally, we also know that the standard normal distribution is symmetric. Thus the fraction of mass under the curve between `-abs(t)` and -infinity (everything to the left of `-abs(t)` under the curve) is exactly the same as the one between `abs(t)` and infinity (everything to the left of `abs(t)` under the curve). Hence we can simply multiply the the value of `pnorm(-abs(t))` by two: `2*pnorm(-abs(t))` for simplicity's sake. Alternatively we could compute the probability for the 'lower tail' and the 'upper tail' separately and build the sum of the two. 

```{r}
# compute the lower tail probability
lowerp <- pnorm(-abs(t))
# compute the upper tail probability
upperp <- pnorm(abs(t), lower.tail = FALSE)
# build the two-tail p-value
twotailp <- lowerp + upperp

# proof that this is the same as above
twotailp == 2*pnorm(-abs(t))

```


### II. Apply the R function for t-tests (i.e., the tl;dr version...)
```{r}
# t-test for H0: mu = 85
t.test(swiss$Fertility, mu = 85)
```

Given this result, we can quite confidently reject H0 and conclude that the population mean of this fertility measure is very unlikely 85. 




## Recap: Hypothesis Tests and T-Statistic 

**What is the t-statistic and how do we compute it?**
```{r}
# load the dataset
data(swiss)

# a) compute sample mean and sample standard deviation, record how many observations we are having
# in our sample, define the population mean (that you want to test for)
sample_mean <- mean(swiss$Fertility)
sample_sd <- sd(swiss$Fertility)
n <- length(swiss$Fertility) # alternatively use nrow(swiss)
mu <- 85

# b) compute the (estimate of the) sample mean standard error
se <- sample_sd / sqrt(n)

# c) compute the t-statistic
t <- (sample_mean - mu) / se
t

# d) check what p-value is associated with that t-statistic 
# i.e., check what fraction of the standard normal distribution has an at least as extreme value as
# the t value we computed.
pval <- 2*pnorm(-abs(t))
pval
```

**Use R's t-statistic function**
Alternatively to the 'manual' approache above, we can use R's `t.test()` function to execute all these steps at once:
```{r}
# t-test for H0: mu = 85
t.test(swiss$Fertility, mu = 85)
```




# References
