# (PART) The Data Pipeline {-}

# Data Sources, Data Gathering, Data Import

This chapter brings together everything you have learned about data formats and structures to tackle a fundamental task: getting data into R. Whether from local files, databases, or the web, importing data correctly is the first step of any analysis.

**Prerequisites:** Understanding of data formats (CSV, XML, JSON, HTML) from Chapters 4-5, and basic R programming from Chapter 2. Familiarity with R data structures (vectors, data frames, lists) is essential.

**What you will learn:**

- How to import data from various file formats (CSV, Excel, JSON, XML)
- How to read data from web sources (JSON, XML, HTML)
- How to scrape data from HTML tables on websites
- Best practices for organizing data import scripts
- How to export data to CSV format

---

In this chapter, we put the key concepts learned so far (text files for data storage, parsers, encoding, data structures) together and apply them to master the first key bottleneck in the data pipeline\index{Data Pipeline}: how to *import* raw data from various sources and *export/store* them for further processing in the pipeline.

## Sources/formats

In the previous chapters, we learned how data is stored in text files and how different data structures/formats/syntaxes help to organize the data in these files. Along the way, we have encountered key data formats that are used in various settings to store and transfer data:

- CSV (typical for rectangular/table-like data)
- Variants of CSV (tab-delimited, fixed-width, etc.)
- XML and JSON (useful for complex/high-dimensional data sets)
- HTML (a markup language to define the structure and layout of webpages)
- Unstructured text

Depending on the *data source*, data might come in one or the other form. With the increasing importance of the Internet as a data source for economic research, properly handling XML, JSON, and HTML is becoming more important. However, in applied economic research, various other formats can be encountered:

- Excel spreadsheets (`.xlsx`, or legacy `.xls`)
- Formats specific to statistical software packages (SPSS: `.sav`, Stata: `.dta`, etc.)
- Built-in R datasets
- Binary formats

While we will cover/revisit how to import all of these formats here, it is important to keep in mind that the learned fundamental concepts are as important (or even more important) than knowing which function to call in R for each of these cases. New formats might evolve and become more relevant in the future for which no R function yet exists. However, the underlying logic of how formats to structure data work will hardly change.

> **In Practice:** Reliable data pipelines are the backbone of reproducible research and trustworthy business analytics. @christensen2018transparency document a "credibility crisis" across the social sciences—many published findings cannot be reproduced, often because data import and cleaning steps were poorly documented or relied on manual processes. The same issues plague business reporting when analyses depend on undocumented manual steps. Best practices include: (1) scripting every step of data acquisition so it can be re-run exactly; (2) storing raw data separately from processed data, never overwriting originals; (3) documenting data sources, access dates, and any manual cleaning decisions; and (4) using checksums to verify file integrity. Academic journals now require replication packages, and businesses increasingly demand auditable data workflows—skills that start with the import scripts covered in this chapter.

## Data gathering procedure

Before we set out to gather/import data from diverse sources, we should start organizing the procedure in an R script. This script will be the beginning of our pipeline!

First, open a new R script in RStudio and save it as `import_data.R` in your `code` folder. Take your time to meaningfully describe what the script is all about in the first few lines of the file:


```{r}
#####################################################
# Data Handling Course: Example Script for
# Data Gathering and Import
#
# Imports data from ...
# Input: links to data sources
# Output: cleaned data as CSV
#
# U. Matter, St.Gallen, 2019
#####################################################

```

RStudio recognizes different sections in a script, whereby section headers are indicated by ending the header text with four or more dashes (`----`). This helps to organize the script into different tasks further. Usually, it makes sense to start with a 'meta' section in which all necessary packages are loaded and fixed variables initialized.


```{r eval=FALSE}
#####################################################
# Data Handling Course: Example Script for
# Data Gathering and Import
#
# Imports data from ...
# Input: links to data sources
# Output: cleaned data as CSV
#
# U. Matter, St.Gallen, 2019
#####################################################


# SET UP --------------
# load packages
library(tidyr)

# set fixed variables
INPUT_PATH <- "/rawdata"
OUTPUT_FILE <- "/final_data/datafile.csv"

```


Finally, we add sections with the actual code (in the case of a data import script, maybe one section per data source).

```{r eval=FALSE}
#####################################################
# Project XY: Data Gathering and Import
#
# First part of the data pipeline of project XY.
# It imports data from ...
# Input: links to data sources
# Output: cleaned data as CSV
#
# U. Matter, St.Gallen, 2019
#####################################################


# SET UP --------------
# load packages
library(tidyr)

# set fixed variables
INPUT_PATH <- "/rawdata"
OUTPUT_FILE <- "/final_data/datafile.csv"


# IMPORT RAW DATA FROM CSVs -------------


```



## Loading/importing rectangular data^[This section is based on @umatter_2018.]

### Loading built-in datasets
We start with the simplest case of loading/importing data. The basic R installation provides some example datasets to try out R's statistics functions. In the introduction to visualization techniques with R and the statistics examples in the chapters to come, we will rely on some of these datasets for simplicity. Note that the usage of these simple datasets shipped with basic R are very helpful when practicing/learning R on your own. Many R packages use these datasets over and over again in their documentation and examples. Moreover, extensive documentations and tutorials online also use these datasets (see for example the [ggplot2 documentation](https://ggplot2.tidyverse.org/)). And, they are very useful when searching help on [Stackoverflow](https://stackoverflow.com/questions/tagged/r) in the context of data analysis/manipulation with R, as you should provide a code example based on some data that everybody can easily load and access.

In order to load such datasets, simply use the `data()`-function:

```{r eval=TRUE}
data(swiss)
```

In this case, we load a dataset called `swiss`. After loading it, the data is stored in a variable of the same name as the dataset (here '`swiss`'). We can inspect it and have a look at the first few rows:

```{r eval=TRUE}
# inspect the structure
str(swiss)

# look at the first few rows
head(swiss)
```

To get a list of all the built-in datasets, type `data()` into the console and hit enter. To get more information about a given dataset, use the help function (e.g., `?swiss`)



### Importing rectangular data from text-files
In most cases of applying R for data analysis, students and researchers rely on importing data from files stored on the hard disk. Typically, such datasets are stored in a text file format such as 'Comma Separated Values' (CSV). In economics, one also frequently encounters data stored in specific formats of commercial statistics/data analysis packages such as SPSS or STATA. Moreover, when collecting data on your own, you might rely on a spreadsheet tool like Microsoft Excel. Data from all these formats can easily be imported into R (in some cases, additional packages have to be loaded, though). Thereby, what happens 'under the hood' is essentially the same for all of these formats. Somebody has implemented the respective *parser* as an R function that accepts a character string with the path or URL to the data source as input.

#### Comma Separated Values (CSV)
Recall how in this format, data values of one observation are stored in one row of a text file, while commas separate the variables/columns. For example, the following code block shows how the first two rows of the `swiss`-dataset would look like when stored in a CSV:

```{}
"District","Fertility","Agriculture","Examination",
    "Education","Catholic","Infant.Mortality"
"Courtelary",80.2,17,15,12,9.96,22.2
```

The function `read.csv()`\index{read.csv()} imports such files from disk into R (in the form of a `data frame`). In this example, the `swiss`-dataset is stored locally on our disk in the folder `data`:


```{r purl=FALSE}
swiss_imported <- read.csv("data/swiss.csv")
```

> **Common Error:** File paths are case-sensitive on Mac/Linux. `"Data/Swiss.csv"` will fail if the actual path is `"data/swiss.csv"`. Also, use forward slashes (`/`) even on Windows—R handles them correctly.

Alternatively, we could use the tidyverse `read_csv()`\index{read\_csv()} function, which is faster, returns a `tibble` instead of a data frame, and does not convert strings to factors by default.

#### Spreadsheets/Excel\index{readxl}
To read excel spreadsheets, we need to install an additional R package called `readxl`.

```{r eval=FALSE}
# install the package
install.packages("readxl")
```

Then we load this additional package ('library') and use the package's `read_excel()`-function to import data from an excel-sheet. In the example below, the same data as above is stored in an excel-sheet called `swiss.xlsx`, again in a folder called `data`.


```{r echo=FALSE, purl=FALSE}
# load the package
library(readxl)

# import data from a spreadsheet
swiss_imported <- read_excel("data/swiss.xlsx")
```


```{r eval=FALSE, warning=FALSE}
# load the package
library(readxl)

# import data from a spreadsheet
swiss_imported <- read_excel("data/swiss.xlsx")
```

#### Data from other data analysis software\index{haven}\index{foreign}
The R packages `foreign` and `haven` contain functions to import data from formats used in other statistics/data analysis software, such as SPSS and STATA.

In the following example we use `haven`'s `read_spss()` function to import a version of the `swiss`-dataset stored in SPSS' `.sav`-format (again stored in the folder called `data`).

```{r echo=FALSE, purl=FALSE, warning=FALSE}
# install the package (if not yet installed):
# install.packages("haven")

# load the package
library(haven)

# read the data
swiss_imported <- read_spss("data/swiss.sav")

```

```{r eval=FALSE}
# install the package (if not yet installed):
# install.packages("haven")

# load the package
library(haven)

# read the data
swiss_imported <- read_spss("data/swiss.sav")

```


## Importing and parsing with `readr`\index{readr}^[This is a summary of the data import chapters in @wickham_cetinkaya_grolemund_2023.]

The `readr` package is automatically installed and loaded with the installation/loading of `tidyverse`. It provides a set of functions to read different types of rectangular data formats and is usually more robust and faster than similar functions in the basic R distribution. Each of these functions expects either a character string with a path pointing to a file or a character string directly containing the data.

### Basic usage of `readr` functions

For example, we can parse the first lines of the swiss dataset directly like this.

```{r}
library(readr)

read_csv(
'"District","Fertility","Agriculture","Examination",
    "Education","Catholic","Infant.Mortality"
"Courtelary",80.2,17,15,12,9.96,22.2')

```

or read the entire `swiss` dataset by pointing to the file
```{r eval=FALSE}
swiss <- read_csv("data/swiss.csv")
```


```{r echo=FALSE, purl=FALSE}
swiss <- read_csv("data/swiss.csv")
```


In either case, the result is a `tibble`:
```{r}
swiss
```

The other `readr` functions have practically the same syntax and behavior. They are used for fixed-width files or CSV-type files with other delimiters than commas.

### Parsing and data types

From inspecting the `swiss` tibble pointed out above, we recognize that `read_csv` not only correctly recognizes observations and columns (parses the CSV correctly) but also automatically guesses the data type of the values in each column. The first column is of type double, the second one of type integer, etc. That is, `read_csv` also parses each column-vector of the data set with the aim of recognizing which data type it is. For example, the data value `"12:00"` could be interpreted simply as a character string. Alternatively, it could also be interpreted as a `time` format.

If `"12:00"` is an element of the vector `c("12:00", "midnight", "noon")` it must be interpreted as a character string. If however it is an element of the vector `c("12:00", "14:30", "20:01")` we probably want R to import this as a `time` format. Now, how can `readr` handle the two cases? In simple terms, the package first guesses for each column vector which type is most appropriate. Then, it uses a couple of lower-level parsing functions (one written for each possible data type) in order to parse each column according to the respective guessed type. We can demonstrate this for the two example vectors above.

```{r}
read_csv('A,B
         12:00, 12:00
         14:30, midnight
         20:01, noon')
```

Under the hood `read_csv()` used the `guess_parser()`- function to determine which type the two vectors likely contain:

```{r}
guess_parser(c("12:00", "midnight", "noon"))
guess_parser(c("12:00", "14:30", "20:01"))
```



## Importing web data formats

### XML in R\index{xml2}^[This section is based on @umatter_2018b.]

There are several XML-parsers already implemented in R packages specifically written for working with XML data. Thus, we do not have to understand the XML syntax in every detail to work with this data format in R. The already familiar package `xml2` (automatically loaded when loading `rvest`) provides the `read_xml()` function which we can use to read the exemplary XML-document.

```{r eval=FALSE}
# load packages
library(xml2)

# parse XML, represent XML document as R object
xml_doc <- read_xml("data/customers.xml")
xml_doc

```


```{r echo=FALSE, purl=FALSE}
# load packages
library(xml2)

# parse XML, represent XML document as R object
xml_doc <- read_xml("data/customers.xml")
xml_doc

```

The same package also has various functions to access, extract, and manipulate data from a parsed XML document. In the following code example, we have a look at the most useful functions for our purposes (see the package's [vignette](https://cran.r-project.org/web/packages/xml2/vignettes/modification.html) for more details).

```{r}
# navigate through the XML document
# (recall the tree-like nested structure similar to HTML)
# navigate downwards
# 'customers' is the root node, persons are ‘their children'
persons <- xml_children(xml_doc)
# navigate sidewards
xml_siblings(persons)
# navigate upwards
xml_parents(persons)

# find data via XPath
customer_names <- xml_find_all(xml_doc, xpath = ".//name")
# extract the data as text
xml_text(customer_names)

```

### JSON in R\index{jsonlite}^[This section is based on @umatter_2018b.]

Again, we can rely on an R package (`jsonlite`) providing high-level functions to read, manipulate, and extract data when working with JSON documents in R. An important difference between working with XML and JSON is that XPath is not compatible with JSON. However, as `jsonlite` represents parsed JSON as R objects of class `list` and/or `data-frame`, we can work with the parsed document as with any other R-object of the same class. The following example illustrates this point.


```{r eval=FALSE}
# load packages
library(jsonlite)

# parse the JSON document shown in the example above
json_doc <- fromJSON("data/person.json")

# look at the structure of the document
str(json_doc)

# navigate the nested lists, extract data
# extract the address part
json_doc$address
# extract the gender (type)
json_doc$gender$type


```


```{r echo=FALSE, purl=FALSE}
# load packages
library(jsonlite)

# parse the JSON document shown in the example above
json_doc <- fromJSON("data/person.json")

# look at the structure of the document
str(json_doc)

# navigate the nested lists, extract data
# extract the address part
json_doc$address
# extract the gender (type)
json_doc$gender$type


```



### Tutorial (advanced): Importing data from a HTML table (on a website)\index{Web Scraping}

> **This tutorial applies:** Chapter 5 (High-Dimensional Data, which covers HTML), Chapter 8.3 (importing web data formats)

In the chapter on high-dimensional data, we discussed the *Hypertext Markup Language (HTML)* as code to define the structure/content of a website and HTML-documents as semi-structured data sources. The following tutorial revisits the basic steps in importing data from an HTML table into R.

The aim of the tutorial is to generate a CSV file containing data on *divided government*\index{Divided Government} in US politics—a situation where different political parties control different branches of government (e.g., one party holds the presidency while another controls one or both chambers of Congress). We use the following Wikipedia page as a data source: https://en.wikipedia.org/wiki/Divided_government_in_the_United_States. The page contains a table indicating the president's party, and the majority party in the US House and the US Senate per Congress (2-year periods). The first few rows of the cleaned data are supposed to look like this:

```{r echo=FALSE}
dgov <- data.table::fread("data/divided_gov.csv")
head(dgov)
```

In a first step, we initialize fixed variables for paths and load additional R packages needed to handle data stored in HTML documents.

```{r eval=FALSE, purl=FALSE}
# SETUP ------------------

# load packages
library(rvest)
library(data.table)

# fixed variables
SOURCE_PATH <- paste0("https://en.wikipedia.org/wiki/",
                      "Divided_government_in_the_United_States")
OUTPUT_PATH <- "data/divided_gov.csv"

```


```{r echo=FALSE, warning=FALSE, message=FALSE}
# SETUP ------------------

# load packages
library(rvest)
library(tidyverse)

# fixed variables
SOURCE_PATH <- paste0("https://en.wikipedia.org/wiki/",
                      "Divided_government_in_the_United_States")
OUTPUT_PATH <- "data/divided_gov.csv"

```

Now we write the part of the script that fetches the data from the Web. This part consists of three steps. First we fetch the entire website (HTML document) from Wikipedia with (`read_html()`). Second, we extract the part of the website containing the table with the data we want (via `html_node()`). Finally, we parse the HTML table and store its content in a data frame called `tab.` The last line of the code chunk below removes the last row of the data frame (you can see on the website that this row is not needed)

```{r}
# FETCH/FORMAT DATA -------------------

# fetch from web
doc <- read_html(SOURCE_PATH)
tab <- html_table(doc,fill=TRUE)[[2]]
tab <- tab[-nrow(tab), ] # remove last row (not containing data)
```

Now we clean the data to get a data set more suitable for data analysis. Note that the original table contains information per congress (2-year periods). However, as the sample above shows, we aim for a panel at the year level. The following code iterates through the rows of the data frame and generates two rows for each congress row (one for each year in the congress).^[See `?strsplit`, `?unlist`, and [this introduction to regular expressions](https://www.oreilly.com/content/an-introduction-to-regular-expressions/) for the background of how this is done in the code example here.]

```{r}
# generate year-level data frame

# prepare loop
all_years <- list() # the container
n <- length(tab$Year) # number of cases to iterate through
length(all_years) <- n
# generate year-level observations. row by row.
for (i in 1:n){
      # select row
      row <- tab[i,]
      y <- row$Year
      #
      years <- unlist(strsplit(x = y, split = "[\\–\\-]", perl = TRUE))
      begin <- as.numeric(years[1])
      end <- as.numeric(years[2])
      tabnew <- data.frame(year = begin:(end-1),
                           president = row$President,
                           senate = row$Senate,
                           house = row$House)
      all_years[[i]] <- tabnew # store in container
}

# stack all rows together
allyears <- bind_rows(all_years)
```

In a last step, we inspect the collected data and write it to a CSV file.

```{r}
# WRITE TO DISK --------------------------------------------------

# inspect
head(allyears)

# write to CSV
write_csv(allyears, file=OUTPUT_PATH)

```

## Key Takeaways

- Organizing your data import process in a well-documented R script is the first step of a reproducible data pipeline.
- R can import data from many sources: built-in datasets, CSV files, Excel spreadsheets, statistical software formats (SPSS, STATA), and web pages.
- The `readr` package provides fast, robust functions for importing rectangular data and automatically guesses column types.
- XML and JSON from web sources can be parsed into R objects using `xml2` and `jsonlite` packages.
- Web scraping with `rvest` allows you to extract data from HTML tables on websites, but the data often requires cleaning before analysis.

> **AI Support Prompt:** If you're having trouble importing a file, describe the issue:
>
> `I'm trying to import a CSV file in R using read_csv(), but [describe problem: 'some columns are wrong type', 'special characters appear garbled', 'the delimiter seems wrong']. Here's what my data looks like: [paste first few lines]. How can I fix this?`
>
> Import issues are common—AI can help diagnose encoding problems, delimiter issues, and type mismatches.

> **AI Support Prompt:** When working with web data or APIs, get help with your approach:
>
> `I want to download data from [describe the source, e.g., 'a government statistics website', 'a REST API that returns JSON']. What's the best approach in R? Should I use httr2, rvest, or something else?`
>
> Different data sources require different tools—AI can help you choose the right one.

> **AI Support Prompt:** To understand web scraping ethics and best practices, ask:
>
> `I want to scrape data from [website]. What should I check before scraping? What are the ethical and legal considerations? How can I be a good citizen when scraping?`
>
> Responsible data collection is important—understanding robots.txt, rate limiting, and terms of service protects you and the websites you access.

## Further Reading

- @wickham_cetinkaya_grolemund_2023: Chapters 7-8 of *R for Data Science* cover data import in detail. Available at https://r4ds.hadley.nz/.
- The `readr` package documentation: https://readr.tidyverse.org/ provides detailed guidance on importing rectangular data.
- The `haven` package documentation: https://haven.tidyverse.org/ covers importing data from SPSS, Stata, and SAS.

## Exercises

1. **Importing CSV files**: Download a CSV dataset from a public source (e.g., World Bank, OECD, Kaggle, or your country's statistics office). Using `read_csv()` from the `readr` package:
   a) Import the data into R
   b) Check the dimensions (`dim()`) and structure (`str()`) of the resulting data frame
   c) Were the column types guessed correctly? If not, specify the correct types using the `col_types` argument.

2. **Excel files**: Create a simple Excel file with at least two sheets, each containing different data (e.g., one with monthly sales, another with product information). Using `readxl`:
   a) List all sheet names in the file
   b) Import each sheet into a separate data frame
   c) Skip the first row if it contains a title rather than column headers

3. **Comparing import functions**: Import the same CSV file using both base R's `read.csv()` and tidyverse's `read_csv()`. Compare:
   a) The time taken for each (use `system.time()`)
   b) The column types assigned
   c) How each handles character strings (check `stringsAsFactors`)

4. **Data from the web**: Write an R script that:
   a) Downloads a JSON file from a public API (e.g., a weather API or government data portal)
   b) Parses it into an R object
   c) Extracts at least three relevant pieces of information








