

# Exploratory Data Analysis

After importing, cleaning, and transforming your data (Chapter 9), the next step is to explore and understand what you have. Exploratory Data Analysis (EDA) is the process of using summary statistics and visualizations to understand the structure, patterns, and peculiarities of your dataset before formal modeling or hypothesis testing.

This chapter covers the essential first steps in any data analysis:

- Computing summary statistics (mean, median, standard deviation)
- Grouping and aggregating data
- Identifying potential issues (outliers, missing values, unexpected patterns)

## Aggregation and summary statistics

Typically, the first step of analyzing a dataset is to get an overview by computing some summary statistics. This helps to better understand the dataset at hand. Key summary statistics of the variables of interest are the mean, standard deviation, median, and number of observations. Together, they give a first idea of how the variables of interest are distributed.

We will continue working with the data prepared in the previous chapter. Let's reload the necessary packages and recreate the dataset:

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)

# recreate the example dataset
df_c <- data.frame(id = c(1:3,1:3),
                   money_spent= c(1000, 2000, 6000, 1500, 3000, 5500),
                   currency = c("CHF", "CHF", "USD", "EUR", "CHF", "USD"),
                   year=c(2017,2017,2017,2018,2018,2018))

df_p <- data.frame(id = 1:4,
                   first_name = c("Anna", "Betty", "Claire", "Diane"),
                   profession = c("Economist", "Data Scientist",
                                  "Data Scientist", "Economist"))

df_merged <- merge(df_p, df_c, by="id")
df_selection <- select(df_merged, id, year, money_spent, currency)

exchange_rates <- data.frame(exchange_rate= c(0.9, 1, 1.2),
                             currency=c("USD", "CHF", "EUR"), stringsAsFactors = FALSE)
df_selection <- merge(df_selection, exchange_rates, by="currency")
df_mutated <- mutate(df_selection, money_spent_chf = money_spent * exchange_rate)
```

R has several built-in functions that help compute summary statistics. In practice, these basic functions are often combined with functions implemented particularly for this step of the analysis, such as `summarise()` provided in `tidyverse`.

As the first output in our report, we want to show the key characteristics of the spending data in one table.

```{r}
summarise(df_mutated,
          mean = mean(money_spent_chf),
          standard_deviation = sd(money_spent_chf),
          median = median(money_spent_chf),
          N = n())
```

### Grouped summaries

Moreover, we can compute the same statistics grouped by certain observation characteristics. For example, we can compute the same summary statistics per year of observation.

```{r message=FALSE, warning=FALSE}
by_year <- group_by(df_mutated, year)
summarise(by_year,
          mean = mean(money_spent_chf),
          standard_deviation = sd(money_spent_chf),
          median = median(money_spent_chf),
          N = n())
```

Grouping by multiple variables is also straightforward:

```{r message=FALSE, warning=FALSE}
by_year_currency <- group_by(df_mutated, year, currency)
summarise(by_year_currency,
          mean = mean(money_spent_chf),
          N = n())
```

### Using apply functions

Alternatively to the more user-friendly (but less flexible) `summarise` function, we can use lower-level functions to compute aggregate statistics provided in the basic R distribution. A good example of such a function is `sapply()`. In simple terms, `sapply()` takes a list as input and applies a function to the content of each element in this list (here: compute a statistic for each column). To illustrate this point, we load the already familiar `swiss` dataset.

```{r}
# load data
data("swiss")
```

Now we want to compute the mean for each variable in this dataset. Technically speaking, a data frame is a list, where each list element is a column of the same length. Thus, we can use `sapply()` to 'apply' the function `mean()` to each of the columns in `swiss`.

```{r}
sapply(swiss, mean)
```

By default, `sapply()` returns a vector or a matrix.^[The related function `lapply()`, returns a list (see `lapply(swiss, mean)`).] We can get a similar result by using `summarise()`. However, we would have to explicitly mention which variables we want as input.

```{r}
summarise(swiss,
          Fertility = mean(Fertility),
          Agriculture = mean(Agriculture)) # etc.
```

## What to look for in your data

When exploring a new dataset, there are several key questions you should ask:

### Check for missing values

Missing values (`NA`) can significantly affect your analysis. Always check how many missing values exist and in which variables:

```{r}
# Count NAs in each column
sapply(df_mutated, function(x) sum(is.na(x)))
```

### Examine the distribution

Look at the range of values to identify potential outliers or data entry errors:

```{r}
summary(swiss)
```

Key things to notice:
- **Min/Max vs Mean/Median**: Large differences between mean and median suggest skewed distributions
- **Extreme values**: Very high max or very low min might indicate outliers
- **Zero or negative values**: In variables where these shouldn't exist (e.g., income, age)

### Check for unexpected values

For categorical variables, examine the unique values:

```{r}
unique(df_mutated$currency)
```

For numeric variables, look at the distribution:

```{r}
# Quick histogram
hist(swiss$Fertility, main = "Distribution of Fertility", xlab = "Fertility")
```

### Identify relationships between variables

A correlation matrix can reveal relationships between numeric variables:

```{r}
# Correlation matrix for swiss data
round(cor(swiss), 2)
```

## From exploration to analysis

The insights gained from EDA should guide your subsequent analysis:

1. **Missing values**: Decide whether to remove observations, impute values, or adjust your analysis approach
2. **Outliers**: Investigate whether they are errors or genuine extreme values
3. **Distributions**: Consider transformations if variables are heavily skewed
4. **Relationships**: Use discovered correlations to inform your modeling choices

The next chapters will cover formal statistical modeling (Chapter 11) and visualization techniques (Chapter 12) that build upon this exploratory foundation.


