
# Basic Statistical Modeling in R

When working with data in economics and social sciences, statistical modeling is often the final destination of the data pipeline. After importing, cleaning, transforming, and exploring your data, you typically want to understand relationships between variables or test specific hypotheses. This chapter introduces the most fundamental tool for this purpose: linear regression using ordinary least squares (OLS).

This chapter covers:

- The concept of statistical modeling
- Using R's `lm()` function for linear regression
- Interpreting regression output
- Understanding statistical significance

## The linear model

Generally, we denote the *dependent variable* (what we want to explain) as $y_i$ and the *explanatory variable* (what we think explains it) as $x_i$. All factors not explicitly included in our model we call the *residuals* or *error term* ($u_{i}$). The basic linear model can be written as:

$y_{i}= \alpha + \beta x_{i} + u_{i}$

where:
- $\alpha$ is the intercept (the expected value of $y$ when $x=0$)
- $\beta$ is the slope coefficient (how much $y$ changes when $x$ increases by one unit)
- $u_{i}$ captures everything else affecting $y$ that we haven't measured

The goal is to estimate $\alpha$ and $\beta$ from our data. The method of *ordinary least squares* finds the values that minimize the sum of squared differences between predicted and actual values of $y$.

### Causality considerations

A key question in applied research is whether we can interpret the estimated relationship as *causal*. The computer scientist [Judea Pearl](https://en.wikipedia.org/wiki/Judea_Pearl) introduced causal diagrams to think about this clearly.

```{r causality, warning=FALSE, message=FALSE, echo=FALSE, out.width="60%", fig.show='hold', fig.align='center', fig.cap= "(ref:causality)", purl=FALSE}
include_graphics("img/causal_diagram.svg")

```
(ref:causality) Causal diagrams. On the left with an endogeneity issue (other factors affect both the dependent and the explanatory variable), and on the right without this endogeneity issue.

For a causal interpretation of $\beta$, the error term $u_{i}$ must be unrelated to $x_i$. Other factors may affect $y$, but they must not simultaneously affect both $y$ and $x$. When this assumption is violated, we have an *endogeneity problem*.

## Regression in R with `lm()`

R provides the `lm()` function (for "linear model") to estimate regression models. Let's use the familiar `swiss` dataset to investigate whether education affects examination success.

```{r message=FALSE, warning=FALSE}
# load the data
data(swiss)
```

The dataset contains information on Swiss provinces. We'll examine whether more years of schooling (`Education`: percentage of draftees with education beyond primary school) is associated with better examination outcomes (`Examination`: percentage receiving the highest mark on standardized tests).

### Visualizing the relationship

Before running a regression, always visualize your data:

```{r}
plot(swiss$Education, swiss$Examination,
     xlab = "Education (% beyond primary)",
     ylab = "Examination (% highest mark)",
     main = "Education vs Examination in Swiss Provinces")
```

There appears to be a positive relationship between education and examination success.

### Fitting a linear model

The `lm()` function takes a formula and a dataset:

```{r}
# fit the model
model <- lm(Examination ~ Education, data = swiss)
```

The formula `Examination ~ Education` means "Examination as a function of Education."

### Viewing the results

The simplest way to see the estimated coefficients:

```{r}
model
```

For more detailed output including statistical tests:

```{r}
summary(model)
```

### Interpreting the output

The `summary()` output contains several important pieces of information:

**Coefficients:**
- `(Intercept)`: The estimated $\hat{\alpha}$. Here, about 5.8, meaning a province with 0% education beyond primary school would be predicted to have about 5.8% achieving the highest examination mark.
- `Education`: The estimated $\hat{\beta}$. Here, about 0.58, meaning each additional percentage point of education beyond primary school is associated with about 0.58 percentage points more draftees achieving the highest mark.

**Statistical significance:**
- `Std. Error`: The uncertainty in our estimate
- `t value`: The ratio of the estimate to its standard error
- `Pr(>|t|)`: The p-value testing whether the true coefficient is zero. Small values (typically < 0.05) suggest the relationship is statistically significant.

The stars (`***`, `**`, `*`) indicate significance levels.

**Model fit:**
- `R-squared`: The proportion of variance in $y$ explained by the model. Here, about 0.33, meaning education explains about 33% of the variation in examination success across provinces.

### Visualizing the fitted model

Add the regression line to your scatter plot:

```{r}
plot(swiss$Education, swiss$Examination,
     xlab = "Education (% beyond primary)",
     ylab = "Examination (% highest mark)")
abline(model, col = "red", lwd = 2)
```

## Multiple regression

Often we want to control for multiple factors. The `lm()` function handles this easily:

```{r}
# include Agriculture as a control variable
model2 <- lm(Examination ~ Education + Agriculture, data = swiss)
summary(model2)
```

The interpretation of `Education` now changes to: the effect of education *holding agriculture constant*.

To include all variables in a dataset:

```{r}
model3 <- lm(Examination ~ ., data = swiss)
summary(model3)
```

The `.` means "all other variables in the dataset."

## Working with model objects

The model object contains useful information you can extract:

```{r}
# get coefficients
coef(model)

# get fitted values (predicted y)
head(fitted(model))

# get residuals (y - predicted y)
head(residuals(model))

# get confidence intervals for coefficients
confint(model)
```

## A word of caution

Statistical significance does not automatically imply causality. In our education example, provinces with higher education levels might differ in other ways that also affect examination success (parental wealth, urbanization, etc.). The simple regression cannot distinguish the effect of education from these confounding factors.

For causal conclusions, you need either:
- A randomized experiment
- A natural experiment or quasi-experimental design
- Careful consideration of all potential confounders

The mathematical details of how OLS estimates are computed, including the derivation of the estimators and a simple implementation, are provided in Appendix B for interested readers.

