---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: 'Lecture 5: Programming with Data'
author: "Prof. Dr. Ulrich Matter"
date: "18/10/2018"
output:
  pdf_document:
    pandoc_args:
    - --filter
    - ../../code/math.py
  html_document:
    highlight: tango
    theme: cerulean
    mathjax: "http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes.css
bibliography: ../references/datahandling.bib

---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
knitr::opts_chunk$set(fig.pos = 'htb!')
```


Programming: Murrell chapter 9.11
- Go through key concepts (maybe extend a bit, take some from cs50)
- Go through the case study (maybe update needed here)




## Working with Data in R: Data Structures and Indeces
(Practical part is important here, they should get an overview, should recall data structures in objects etc.)

### Vectors and Lists
```{r}
# A vector containing numeric (or integer) values
numeric_vector <- 10:20
numeric_vector[2]
numeric_vector[2:5]

# A string vector ('a vector containing text')
string_vector <- c("a", "b", "c")
string_vector[-3]

# Lists
# A list can contain different types of elements, for example a numeric vector and a string_vector
mylist <- list(numbers = numeric_vector, letters = string_vector)
mylist

# We can access the elements of a list in various ways
# with the element's name
mylist$numbers
mylist["numbers"]
# via the index
mylist[1]
# with [[]] we can access directly the content of the element
mylist[[1]]

# lists can also be nested (list of lists of lists....)
mynestedlist <- list(a = mylist, b = 1:5)
```

### Matrices and Data Frames
```{r}

# matrices
mymatrix <- matrix(numeric_vector, nrow = 4)
# get the second row
mymatrix[2,]
# get the first two columns
mymatrix[, 1:2]

# data frames ("lists as columns")
mydf <- data.frame(Name = c("Alice", "Betty", "Claire"), Age = c(20, 30, 45))
mydf
# select the age column
mydf$Age
mydf[, "Age"]
mydf[, 2]
# select the second row
mydf[2,]


```

### Classes and Data Structure
```{r}
# have a look at what kind of object you are dealing with
class(mydf)
class(mymatrix)

# have a closer look at the data structure
str(mydf)
```


# References
